---
title: "ESS for multi-groups borrowing with Hierarchical Models"
output: html_vignette
vignette: >
  %\VignetteIndexEntry{ESS for multi-groups borrowing with Hierarchical Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)

set.seed(20260218)

library(geoess)
library(rstan)

rstan::rstan_options(auto_write = FALSE)
options(mc.cores = 2)
```

## Introduction

This vignette demonstrates geometric effective sample size (ESS) for
multi-group borrowing in a Normal--Normal hierarchical model. We focus on:

- local block ESS for hyperparameters \(\phi = (\mu, \tau^2)\),
- local conditional ESS for group effects \(\theta_i \mid \phi\) on both the
  per-observation and per-group scales, and
- global ESS under prior and posterior weighting distributions.

The workflow follows geometric definitions in `geoess`, where local ESS is
curvature-based and global ESS is an average of local ESS under a chosen
weighting distribution \(Q\).

## Model

We use the hierarchical model
\[
y_{ij} \mid \theta_i, \sigma_i^2 \sim \mathcal{N}(\theta_i, \sigma_i^2),
\qquad i=1,\dots,K,
\]
\[
\theta_i \mid \mu, \tau^2 \sim \mathcal{N}(\mu, \tau^2),
\qquad
\mu \sim \mathcal{N}(m_0, s_0^2),
\qquad
\tau^2 \sim \operatorname{InvGamma}(a_\tau, b_\tau).
\]

Within-group variances are treated by plug-in estimates,
\[
\widehat{\sigma}_i^2 = \operatorname{var}(y_i),
\]
which are used as known inputs in likelihood-level unit information for
\(\theta_i\).

For hyperparameters \(\phi = (\mu, \tau^2)\), per-group unit information is
\[
I_{1,\phi}(\tau^2)
=
\begin{pmatrix}
1/\tau^2 & 0 \\
0 & 1/(2\tau^4)
\end{pmatrix}.
\]

## Simulation Design

```{r simulation-design}
# Baseline simulation settings
K <- 10
# Use heterogeneous group sizes to reflect a more realistic multi-group study.
set.seed(20260217)
n_i <- sample(c(3, 5, 8, 12, 18, 24, 30, 35, 45, 60), size = K, replace = FALSE)
mu0 <- 0
tau0 <- 0.5
tau0_sq <- tau0^2

# Hyperprior settings for fitting and ESS calculations
m0 <- 0
s0 <- 1
a_tau <- 3
b_tau <- 0.5

simulate_hier_data <- function(K, n_i, mu0, tau0, seed = 1) {
  set.seed(seed)
  if (length(n_i) == 1L) n_i <- rep(n_i, K)

  # Heterogeneous within-group SDs
  sigma_true <- sample(c(0.5, 1, 2), size = K, replace = TRUE)
  sigma_true2 <- sigma_true^2

  theta_true <- rnorm(K, mean = mu0, sd = tau0)
  y_list <- lapply(seq_len(K), function(i) {
    rnorm(n_i[i], mean = theta_true[i], sd = sigma_true[i])
  })

  sigma_hat2 <- vapply(y_list, var, numeric(1))

  list(
    K = K,
    n_i = n_i,
    mu0 = mu0,
    tau0 = tau0,
    tau0_sq = tau0^2,
    theta_true = theta_true,
    sigma_true2 = sigma_true2,
    y_list = y_list,
    sigma_hat2 = sigma_hat2
  )
}

sim <- simulate_hier_data(K = K, n_i = n_i, mu0 = mu0, tau0 = tau0, seed = 20260218)

data.frame(
  K = sim$K,
  n_min = min(sim$n_i),
  n_mean = mean(sim$n_i),
  n_max = max(sim$n_i),
  n_i = paste(sim$n_i, collapse = ", "),
  true_mu0 = sim$mu0,
  true_tau0_sq = sim$tau0_sq,
  mean_sigma_true_sq = mean(sim$sigma_true2),
  mean_sigma_hat_sq = mean(sim$sigma_hat2)
)
```

The baseline simulation uses \(K=10\) groups with heterogeneous group sizes
\(\{n_1,\dots,n_{10}\} = \{\) `r paste(sim$n_i, collapse = ", ")` \(\}\), true
\(\mu_0 = 0\), true \(\tau_0^2 = 0.25\), and hyperparameters
\(m_0 = 0\), \(s_0 = 1\), \(a_\tau = 3\), and \(b_\tau = 0.5\). The plug-in
within-group variances \(\widehat{\sigma}_i^2\) range from
`r format(min(sim$sigma_hat2), digits = 3)` to
`r format(max(sim$sigma_hat2), digits = 3)`, with mean
`r format(mean(sim$sigma_hat2), digits = 3)`.

## Stan Implementation

```{r stan-implementation}
hier_stan_code <- "
data {
  int<lower=1> N;
  int<lower=1> K;
  array[N] int<lower=1, upper=K> group;
  vector[N] y;
  vector<lower=0>[K] sigma_hat;
  real m0;
  real<lower=0> s0;
  real<lower=0> a_tau;
  real<lower=0> b_tau;
}
parameters {
  vector[K] theta;
  real mu;
  real<lower=0> tau2;
}
model {
  mu ~ normal(m0, s0);
  tau2 ~ inv_gamma(a_tau, b_tau);
  theta ~ normal(mu, sqrt(tau2));

  for (n in 1:N) {
    y[n] ~ normal(theta[group[n]], sigma_hat[group[n]]);
  }
}
"

make_stan_data <- function(sim, m0, s0, a_tau, b_tau) {
  y <- unlist(sim$y_list)
  group <- rep(seq_len(sim$K), times = sim$n_i)
  list(
    N = length(y),
    K = sim$K,
    group = as.integer(group),
    y = as.numeric(y),
    sigma_hat = sqrt(sim$sigma_hat2),
    m0 = m0,
    s0 = s0,
    a_tau = a_tau,
    b_tau = b_tau
  )
}

stan_model_hier <- rstan::stan_model(model_code = hier_stan_code)
stan_data <- make_stan_data(sim, m0, s0, a_tau, b_tau)

fit_hier <- rstan::sampling(
  object = stan_model_hier,
  data = stan_data,
  chains = 2,
  iter = 700,
  warmup = 350,
  seed = 20260218,
  refresh = 0,
  control = list(adapt_delta = 0.9)
)

theta_draws <- geoess_draws(fit_hier, pars = "theta", as = "matrix")
mu_draws <- as.numeric(geoess_draws(fit_hier, pars = "mu", as = "matrix"))
tau2_draws <- as.numeric(geoess_draws(fit_hier, pars = "tau2", as = "matrix"))

theta_pars <- colnames(theta_draws)
phi_pars <- c("mu", "tau2")

post_draws <- cbind(theta_draws, mu = mu_draws, tau2 = tau2_draws)

# Prior draws for prior-averaged global ESS
n_prior <- nrow(post_draws)
set.seed(20260219)
mu_prior <- rnorm(n_prior, mean = m0, sd = s0)
tau2_prior <- 1 / rgamma(n_prior, shape = a_tau, rate = b_tau)
theta_prior <- t(vapply(seq_len(n_prior), function(s) {
  rnorm(sim$K, mean = mu_prior[s], sd = sqrt(tau2_prior[s]))
}, numeric(sim$K)))

colnames(theta_prior) <- theta_pars
prior_draws <- cbind(theta_prior, mu = mu_prior, tau2 = tau2_prior)

summary(tau2_draws)
```

For more accurate posterior/global ESS summaries, increase `iter`, `warmup`,
and number of chains.

## Local ESS: Block vs Conditional

```{r local-ess-helpers}
build_full_matrix <- function(M_theta, M_thetaphi, M_phi) {
  rbind(
    cbind(M_theta, M_thetaphi),
    cbind(t(M_thetaphi), M_phi)
  )
}

split_full_matrix <- function(M_full, K) {
  idx_theta <- seq_len(K)
  idx_phi <- (K + 1):(K + 2)
  list(
    Htt = M_full[idx_theta, idx_theta, drop = FALSE],
    Htp = M_full[idx_theta, idx_phi, drop = FALSE],
    Hpp = M_full[idx_phi, idx_phi, drop = FALSE]
  )
}

clip_psd <- function(M) {
  M <- 0.5 * (M + t(M))
  ev <- eigen(M, symmetric = TRUE)
  ev$values[ev$values < 0] <- 0
  ev$vectors %*% diag(ev$values, nrow = length(ev$values)) %*% t(ev$vectors)
}

prior_curvature_blocks <- function(theta, mu, tau2, m0, s0, a_tau, b_tau) {
  K <- length(theta)
  delta <- theta - mu
  s1 <- sum(delta)
  s2 <- sum(delta^2)

  Htt <- diag(1 / tau2, K)
  Htp <- cbind(
    mu = rep(-1 / tau2, K),
    tau2 = -(delta) / (tau2^2)
  )

  Hpp11 <- K / tau2 + 1 / (s0^2)
  Hpp12 <- s1 / (tau2^2)
  Hpp22 <- (s2 + 2 * b_tau) / (tau2^3) - (K / 2 + a_tau + 1) / (tau2^2)
  Hpp <- matrix(c(Hpp11, Hpp12, Hpp12, Hpp22), nrow = 2, byrow = TRUE)
  Hpp <- 0.5 * (Hpp + t(Hpp))

  list(Htt = Htt, Htp = Htp, Hpp = Hpp)
}

unit_info_phi <- function(tau2) {
  matrix(c(
    1 / tau2, 0,
    0, 1 / (2 * tau2^2)
  ), nrow = 2, byrow = TRUE)
}

local_ess_summary <- function(theta_ref, mu_ref, tau2_ref, sigma_hat2, n_i,
                              m0, s0, a_tau, b_tau) {
  K <- length(theta_ref)

  blocks <- prior_curvature_blocks(
    theta = theta_ref,
    mu = mu_ref,
    tau2 = tau2_ref,
    m0 = m0,
    s0 = s0,
    a_tau = a_tau,
    b_tau = b_tau
  )

  I1_theta <- diag(1 / sigma_hat2, K)
  I1_phi <- unit_info_phi(tau2_ref)

  I1_full <- build_full_matrix(I1_theta, matrix(0, nrow = K, ncol = 2), I1_phi)
  Ipi_full <- build_full_matrix(blocks$Htt, blocks$Htp, blocks$Hpp)
  Ipi_full_clip <- clip_psd(Ipi_full)
  blocks_clip <- split_full_matrix(Ipi_full_clip, K)

  idx_phi <- (K + 1):(K + 2)
  ess_theta_cond <- vapply(seq_len(K), function(i) {
    geoess_conditional(
      I1 = I1_full,
      Ipi = Ipi_full,
      idx_a = i,
      idx_b = idx_phi,
      method = "scalar"
    )$ess_cond
  }, numeric(1))
  ess_theta_cond_clip <- vapply(seq_len(K), function(i) {
    geoess_conditional(
      I1 = I1_full,
      Ipi = Ipi_full_clip,
      idx_a = i,
      idx_b = idx_phi,
      method = "scalar"
    )$ess_cond
  }, numeric(1))
  ess_theta_cond_per_group <- ess_theta_cond / n_i
  ess_theta_cond_per_group_clip <- ess_theta_cond_clip / n_i

  ess_phi_block <- geoess_trace(I1_phi, blocks$Hpp)
  ess_phi_block_clip <- geoess_trace(I1_phi, blocks_clip$Hpp)
  ess_mu <- blocks$Hpp[1, 1] / I1_phi[1, 1]
  ess_mu_clip <- blocks_clip$Hpp[1, 1] / I1_phi[1, 1]
  ess_tau2 <- blocks$Hpp[2, 2] / I1_phi[2, 2]
  ess_tau2_clip <- blocks_clip$Hpp[2, 2] / I1_phi[2, 2]

  list(
    ess_phi_block = ess_phi_block,
    ess_phi_block_clip = ess_phi_block_clip,
    ess_mu = ess_mu,
    ess_mu_clip = ess_mu_clip,
    ess_tau2 = ess_tau2,
    ess_tau2_clip = ess_tau2_clip,
    ess_theta_cond_per_obs = ess_theta_cond,
    ess_theta_cond_per_obs_clip = ess_theta_cond_clip,
    ess_theta_cond_per_group = ess_theta_cond_per_group,
    ess_theta_cond_per_group_clip = ess_theta_cond_per_group_clip,
    I1_phi = I1_phi,
    Ipi_phi = blocks$Hpp,
    Ipi_phi_clip = blocks_clip$Hpp
  )
}
```

```{r local-ess-computation}
# Reference points
mu_ref_prior <- m0
tau2_ref_prior <- b_tau / (a_tau + 1)  # prior mode for Inv-Gamma(a_tau, b_tau)
theta_ref_prior <- rep(mu_ref_prior, sim$K)

mu_ref_post <- mean(mu_draws)
tau2_ref_post <- mean(tau2_draws)
theta_ref_post <- colMeans(theta_draws)

loc_prior <- local_ess_summary(
  theta_ref = theta_ref_prior,
  mu_ref = mu_ref_prior,
  tau2_ref = tau2_ref_prior,
  sigma_hat2 = sim$sigma_hat2,
  n_i = sim$n_i,
  m0 = m0,
  s0 = s0,
  a_tau = a_tau,
  b_tau = b_tau
)

loc_post <- local_ess_summary(
  theta_ref = theta_ref_post,
  mu_ref = mu_ref_post,
  tau2_ref = tau2_ref_post,
  sigma_hat2 = sim$sigma_hat2,
  n_i = sim$n_i,
  m0 = m0,
  s0 = s0,
  a_tau = a_tau,
  b_tau = b_tau
)

local_table <- data.frame(
  reference = c("Prior reference", "Posterior mean reference"),
  ESS_phi_block_per_group = c(loc_prior$ess_phi_block, loc_post$ess_phi_block),
  ESS_phi_block_per_group_clip = c(loc_prior$ess_phi_block_clip, loc_post$ess_phi_block_clip),
  ESS_mu_per_group = c(loc_prior$ess_mu, loc_post$ess_mu),
  ESS_mu_per_group_clip = c(loc_prior$ess_mu_clip, loc_post$ess_mu_clip),
  ESS_tau2_per_group = c(loc_prior$ess_tau2, loc_post$ess_tau2),
  ESS_tau2_per_group_clip = c(loc_prior$ess_tau2_clip, loc_post$ess_tau2_clip),
  mean_ESS_theta_cond_per_obs = c(mean(loc_prior$ess_theta_cond_per_obs), mean(loc_post$ess_theta_cond_per_obs)),
  mean_ESS_theta_cond_per_obs_clip = c(mean(loc_prior$ess_theta_cond_per_obs_clip), mean(loc_post$ess_theta_cond_per_obs_clip)),
  median_ESS_theta_cond_per_obs = c(median(loc_prior$ess_theta_cond_per_obs), median(loc_post$ess_theta_cond_per_obs)),
  median_ESS_theta_cond_per_obs_clip = c(median(loc_prior$ess_theta_cond_per_obs_clip), median(loc_post$ess_theta_cond_per_obs_clip)),
  mean_ESS_theta_cond_per_group = c(mean(loc_prior$ess_theta_cond_per_group), mean(loc_post$ess_theta_cond_per_group)),
  mean_ESS_theta_cond_per_group_clip = c(mean(loc_prior$ess_theta_cond_per_group_clip), mean(loc_post$ess_theta_cond_per_group_clip)),
  median_ESS_theta_cond_per_group = c(median(loc_prior$ess_theta_cond_per_group), median(loc_post$ess_theta_cond_per_group)),
  median_ESS_theta_cond_per_group_clip = c(median(loc_prior$ess_theta_cond_per_group_clip), median(loc_post$ess_theta_cond_per_group_clip))
)

local_group_table <- data.frame(
  group = seq_len(sim$K),
  n_i = sim$n_i,
  sigma_hat2 = sim$sigma_hat2,
  ESS_local_prior_ref_per_obs = loc_prior$ess_theta_cond_per_obs,
  ESS_local_prior_ref_per_obs_clip = loc_prior$ess_theta_cond_per_obs_clip,
  ESS_local_post_ref_per_obs = loc_post$ess_theta_cond_per_obs,
  ESS_local_post_ref_per_obs_clip = loc_post$ess_theta_cond_per_obs_clip,
  ESS_local_prior_ref_per_group = loc_prior$ess_theta_cond_per_group,
  ESS_local_prior_ref_per_group_clip = loc_prior$ess_theta_cond_per_group_clip,
  ESS_local_post_ref_per_group = loc_post$ess_theta_cond_per_group,
  ESS_local_post_ref_per_group_clip = loc_post$ess_theta_cond_per_group_clip
)

summarize_group_ess <- function(x) {
  c(
    mean = mean(x),
    median = median(x),
    q10 = as.numeric(stats::quantile(x, probs = 0.10)),
    q90 = as.numeric(stats::quantile(x, probs = 0.90))
  )
}

local_group_summary <- data.frame(
  reference = c(
    "Local prior reference per observation (signed)",
    "Local posterior reference per observation (signed)",
    "Local prior reference per observation (clipped)",
    "Local posterior reference per observation (clipped)",
    "Local prior reference per group (signed)",
    "Local posterior reference per group (signed)",
    "Local prior reference per group (clipped)",
    "Local posterior reference per group (clipped)"
  ),
  rbind(
    summarize_group_ess(loc_prior$ess_theta_cond_per_obs),
    summarize_group_ess(loc_post$ess_theta_cond_per_obs),
    summarize_group_ess(loc_prior$ess_theta_cond_per_obs_clip),
    summarize_group_ess(loc_post$ess_theta_cond_per_obs_clip),
    summarize_group_ess(loc_prior$ess_theta_cond_per_group),
    summarize_group_ess(loc_post$ess_theta_cond_per_group),
    summarize_group_ess(loc_prior$ess_theta_cond_per_group_clip),
    summarize_group_ess(loc_post$ess_theta_cond_per_group_clip)
  ),
  row.names = NULL
)
invisible(NULL)
```

```{r local-ess-plot}
plot(
  sim$sigma_hat2,
  loc_post$ess_theta_cond_per_obs,
  pch = 19,
  col = "steelblue",
  xlab = expression(hat(sigma)[i]^2),
  ylab = expression(ESS[theta[i] ~ "|" ~ phi]^local ~ "(per observation)"),
  main = expression("Local Per-Observation ESS at Posterior Reference vs " * hat(sigma)[i]^2)
)
text(sim$sigma_hat2, loc_post$ess_theta_cond_per_obs, labels = seq_len(sim$K), pos = 3, cex = 0.75)
```

In this model, \(I_{1,\theta_i\mid\phi} = I_{1,\theta_i} = 1/\widehat{\sigma}_i^2\)
because observation-level likelihood terms involve \(\theta_i\) but not
\(\phi=(\mu,\tau^2)\), so \(I_{1,\theta\phi}=0\).

```{r anisotropy-phi}
phi_spectrum_post <- geoess_spectrum(loc_post$I1_phi, loc_post$Ipi_phi)

data.frame(
  direction = c("eig1", "eig2"),
  value = phi_spectrum_post$values
)
```

## Global ESS: Prior- vs Posterior-Averaged

```{r global-ess-helpers}
make_ipi_provider <- function(K, m0, s0, a_tau, b_tau, clip = FALSE) {
  force(K); force(m0); force(s0); force(a_tau); force(b_tau)
  force(clip)
  function(Theta, ...) {
    theta <- Theta[seq_len(K)]
    mu <- Theta[K + 1]
    tau2 <- Theta[K + 2]
    blocks <- prior_curvature_blocks(theta, mu, tau2, m0, s0, a_tau, b_tau)
    if (isTRUE(clip)) {
      Ipi_full <- build_full_matrix(blocks$Htt, blocks$Htp, blocks$Hpp)
      Ipi_full <- clip_psd(Ipi_full)
      return(list(Ipi_full = Ipi_full))
    }
    list(
      Ipi_thetatheta = blocks$Htt,
      Ipi_thetaphi = blocks$Htp,
      Ipi_phiphi = blocks$Hpp
    )
  }
}

make_i1_provider <- function(K, sigma_hat2) {
  force(K); force(sigma_hat2)
  function(Theta, ...) {
    tau2 <- Theta[K + 2]
    list(
      I1_thetatheta = diag(1 / sigma_hat2, K),
      I1_thetaphi = matrix(0, nrow = K, ncol = 2),
      I1_phiphi = unit_info_phi(tau2)
    )
  }
}

ipi_provider <- make_ipi_provider(sim$K, m0, s0, a_tau, b_tau, clip = FALSE)
ipi_provider_clip <- make_ipi_provider(sim$K, m0, s0, a_tau, b_tau, clip = TRUE)
i1_provider <- make_i1_provider(sim$K, sim$sigma_hat2)
```

```{r global-ess-computation}
res_global_prior <- geoess_hier_global(
  Q = "custom",
  draws = prior_draws,
  pars_theta = theta_pars,
  pars_phi = phi_pars,
  group_index = seq_len(sim$K),
  Ipi_provider = ipi_provider,
  I1_provider = i1_provider,
  return_by_group = TRUE
)

res_global_prior_clip <- geoess_hier_global(
  Q = "custom",
  draws = prior_draws,
  pars_theta = theta_pars,
  pars_phi = phi_pars,
  group_index = seq_len(sim$K),
  Ipi_provider = ipi_provider_clip,
  I1_provider = i1_provider,
  return_by_group = TRUE
)

res_global_post <- geoess_hier_global(
  Q = "custom",
  draws = post_draws,
  pars_theta = theta_pars,
  pars_phi = phi_pars,
  group_index = seq_len(sim$K),
  Ipi_provider = ipi_provider,
  I1_provider = i1_provider,
  return_by_group = TRUE
)

res_global_post_clip <- geoess_hier_global(
  Q = "custom",
  draws = post_draws,
  pars_theta = theta_pars,
  pars_phi = phi_pars,
  group_index = seq_len(sim$K),
  Ipi_provider = ipi_provider_clip,
  I1_provider = i1_provider,
  return_by_group = TRUE
)

global_table <- data.frame(
  weighting_distribution = c("Prior (Q = pi(Theta))", "Posterior (Q = pi(Theta | y))"),
  global_ESS_phi_block_per_group = c(
    res_global_prior$ess_phi_global$estimate,
    res_global_post$ess_phi_global$estimate
  ),
  global_ESS_phi_block_per_group_clip = c(
    res_global_prior_clip$ess_phi_global$estimate,
    res_global_post_clip$ess_phi_global$estimate
  ),
  MCSE_phi_block = c(
    res_global_prior$ess_phi_global$mcse,
    res_global_post$ess_phi_global$mcse
  ),
  MCSE_phi_block_clip = c(
    res_global_prior_clip$ess_phi_global$mcse,
    res_global_post_clip$ess_phi_global$mcse
  ),
  mean_ESS_theta_cond_per_obs = c(
    mean(res_global_prior$ess_theta_cond_global),
    mean(res_global_post$ess_theta_cond_global)
  ),
  mean_ESS_theta_cond_per_obs_clip = c(
    mean(res_global_prior_clip$ess_theta_cond_global),
    mean(res_global_post_clip$ess_theta_cond_global)
  ),
  median_ESS_theta_cond_per_obs = c(
    median(res_global_prior$ess_theta_cond_global),
    median(res_global_post$ess_theta_cond_global)
  ),
  median_ESS_theta_cond_per_obs_clip = c(
    median(res_global_prior_clip$ess_theta_cond_global),
    median(res_global_post_clip$ess_theta_cond_global)
  ),
  mean_ESS_theta_cond_per_group = c(
    mean(res_global_prior$ess_theta_cond_global / sim$n_i),
    mean(res_global_post$ess_theta_cond_global / sim$n_i)
  ),
  mean_ESS_theta_cond_per_group_clip = c(
    mean(res_global_prior_clip$ess_theta_cond_global / sim$n_i),
    mean(res_global_post_clip$ess_theta_cond_global / sim$n_i)
  ),
  median_ESS_theta_cond_per_group = c(
    median(res_global_prior$ess_theta_cond_global / sim$n_i),
    median(res_global_post$ess_theta_cond_global / sim$n_i)
  ),
  median_ESS_theta_cond_per_group_clip = c(
    median(res_global_prior_clip$ess_theta_cond_global / sim$n_i),
    median(res_global_post_clip$ess_theta_cond_global / sim$n_i)
  )
)

global_group_post_table <- data.frame(
  group = seq_len(sim$K),
  n_i = sim$n_i,
  sigma_hat2 = sim$sigma_hat2,
  ESS_global_post_per_obs = res_global_post$ess_theta_cond_global,
  ESS_global_post_per_obs_clip = res_global_post_clip$ess_theta_cond_global,
  ESS_global_post_per_group = res_global_post$ess_theta_cond_global / sim$n_i,
  ESS_global_post_per_group_clip = res_global_post_clip$ess_theta_cond_global / sim$n_i,
  MCSE_per_obs = res_global_post$ess_theta_mcse,
  MCSE_per_obs_clip = res_global_post_clip$ess_theta_mcse,
  MCSE_per_group = res_global_post$ess_theta_mcse / sim$n_i,
  MCSE_per_group_clip = res_global_post_clip$ess_theta_mcse / sim$n_i
)

global_group_prior_table <- data.frame(
  group = seq_len(sim$K),
  n_i = sim$n_i,
  ESS_global_prior_per_obs = res_global_prior$ess_theta_cond_global,
  ESS_global_prior_per_obs_clip = res_global_prior_clip$ess_theta_cond_global,
  ESS_global_prior_per_group = res_global_prior$ess_theta_cond_global / sim$n_i,
  ESS_global_prior_per_group_clip = res_global_prior_clip$ess_theta_cond_global / sim$n_i,
  MCSE_per_obs = res_global_prior$ess_theta_mcse,
  MCSE_per_obs_clip = res_global_prior_clip$ess_theta_mcse,
  MCSE_per_group = res_global_prior$ess_theta_mcse / sim$n_i,
  MCSE_per_group_clip = res_global_prior_clip$ess_theta_mcse / sim$n_i
)

global_group_summary <- data.frame(
  source = c(
    "Global prior per observation (signed)",
    "Global posterior per observation (signed)",
    "Global prior per observation (clipped)",
    "Global posterior per observation (clipped)",
    "Global prior per group (signed)",
    "Global posterior per group (signed)",
    "Global prior per group (clipped)",
    "Global posterior per group (clipped)"
  ),
  rbind(
    summarize_group_ess(res_global_prior$ess_theta_cond_global),
    summarize_group_ess(res_global_post$ess_theta_cond_global),
    summarize_group_ess(res_global_prior_clip$ess_theta_cond_global),
    summarize_group_ess(res_global_post_clip$ess_theta_cond_global),
    summarize_group_ess(res_global_prior$ess_theta_cond_global / sim$n_i),
    summarize_group_ess(res_global_post$ess_theta_cond_global / sim$n_i),
    summarize_group_ess(res_global_prior_clip$ess_theta_cond_global / sim$n_i),
    summarize_group_ess(res_global_post_clip$ess_theta_cond_global / sim$n_i)
  ),
  row.names = NULL
)
invisible(NULL)
```

`geoess_hier_global` computes local conditional ESS draw-by-draw and then
averages under \(Q\). The underlying objects retained in this vignette include
both signed ESS and clipped ESS (negative prior-curvature eigenvalues set to
zero) for the same group-specific global conditional quantity
\(\mathrm{ESS}^{glob}_{\theta_i\mid\phi}\), reported on both the
per-observation and per-group scales, with Monte Carlo SE from finite draws.

## Hierarchical ESS Summary

Because hierarchical ESS is both level- and unit-specific, we report three
distinct quantities:

- the block ESS for the hyperparameters \(\phi=(\mu,\tau^2)\), measured in
  group units;
- the observation-specific conditional ESS for \(\theta_i \mid \phi\); and
- the group-specific conditional ESS for \(\theta_i \mid \phi\), obtained by
  scaling the observation-specific quantity by \(1/n_i\).

The display tables below use the clipped ESS summaries, which match the
nonnegative reporting convention used in the manuscript. The full signed
objects remain available as `local_table`, `global_table`,
`local_group_table`, and the corresponding global group tables.

```{r hierarchical-ess-display}
hier_main_summary_display <- data.frame(
  Parameter = c(
    "$\\phi = (\\mu, \\tau^2)$",
    "obs-spec $\\theta_i \\mid \\phi$, average",
    "grp-spec $\\theta_i \\mid \\phi$, average"
  ),
  `Local ESS<br>Prior mean` = c(
    loc_prior$ess_phi_block_clip,
    mean(loc_prior$ess_theta_cond_per_obs_clip),
    mean(loc_prior$ess_theta_cond_per_group_clip)
  ),
  `Local ESS<br>Posterior mean` = c(
    loc_post$ess_phi_block_clip,
    mean(loc_post$ess_theta_cond_per_obs_clip),
    mean(loc_post$ess_theta_cond_per_group_clip)
  ),
  `Global ESS<br>Prior average` = c(
    res_global_prior_clip$ess_phi_global$estimate,
    mean(res_global_prior_clip$ess_theta_cond_global),
    mean(res_global_prior_clip$ess_theta_cond_global / sim$n_i)
  ),
  `Global ESS<br>Posterior average` = c(
    res_global_post_clip$ess_phi_global$estimate,
    mean(res_global_post_clip$ess_theta_cond_global),
    mean(res_global_post_clip$ess_theta_cond_global / sim$n_i)
  ),
  check.names = FALSE
)

hier_main_summary_display[-1] <- lapply(hier_main_summary_display[-1], round, 1)

knitr::kable(
  hier_main_summary_display,
  digits = 1,
  escape = FALSE,
  caption = paste(
    "Hierarchical ESS summaries for the simulation.",
    "This table reports clipped ESS for the hyperparameter block",
    "$\\phi=(\\mu,\\tau^2)$, alongside the average observation-specific",
    "and group-specific conditional ESS for $\\theta_i \\mid \\phi$."
  )
)

hier_group_specific_display <- data.frame(
  Group = seq_len(sim$K),
  `$n_i$` = sim$n_i,
  `$\\hat{\\sigma}_i^2$` = round(sim$sigma_hat2, 2),
  `Local ESS<br>Prior mean<br>obs-spec` = round(loc_prior$ess_theta_cond_per_obs_clip, 1),
  `Local ESS<br>Prior mean<br>grp-spec` = round(loc_prior$ess_theta_cond_per_group_clip, 1),
  `Local ESS<br>Posterior mean<br>obs-spec` = round(loc_post$ess_theta_cond_per_obs_clip, 1),
  `Local ESS<br>Posterior mean<br>grp-spec` = round(loc_post$ess_theta_cond_per_group_clip, 1),
  `Global ESS<br>Prior average<br>obs-spec` = round(res_global_prior_clip$ess_theta_cond_global, 1),
  `Global ESS<br>Prior average<br>grp-spec` = round(res_global_prior_clip$ess_theta_cond_global / sim$n_i, 1),
  `Global ESS<br>Posterior average<br>obs-spec` = round(res_global_post_clip$ess_theta_cond_global, 1),
  `Global ESS<br>Posterior average<br>grp-spec` = round(res_global_post_clip$ess_theta_cond_global / sim$n_i, 1),
  check.names = FALSE
)

knitr::kable(
  hier_group_specific_display,
  digits = 1,
  escape = FALSE,
  caption = paste(
    "Group-wise observation-specific and group-specific conditional ESS values.",
    "Local ESS is evaluated at the prior and posterior means, while global ESS",
    "is averaged over the prior and posterior distributions."
  )
)
```

The summary table highlights three consistent patterns. First, local and global
ESS can diverge materially. For example, the average observation-specific
conditional ESS drops from 18.9 at the prior mean to 9.5 at the posterior mean,
but the corresponding global summaries are more stable at 12.5 under prior
averaging and 10.9 under posterior averaging. Second, local conditional ESS is
more sensitive to the reference point than global ESS is to the weighting
distribution. Third, the hyperparameter block ESS is comparatively stable once
averaged globally, with both prior- and posterior-averaged values near 12.1
groups.

The group-wise results make the same point more sharply. Groups 2 and 5 show an
extreme divergence between local and global ESS, and that divergence comes
entirely from the conditional prior curvature. For Group 2, the conditional
curvature is very weak at the chosen reference point, so the local ESS is near
zero, but it is much larger on average over the hierarchical uncertainty,
yielding global group-specific ESS of 7.2 under prior averaging and 6.4 under
posterior averaging. Group 5 shows the opposite pattern: the reference point
falls in a region of unusually strong local curvature, producing large local
ESS, but the broader prior and posterior averages are much smaller. Averaging
therefore dampens unusually strong or weak local curvature and provides a more
stable summary of borrowing.

There is a common intuition that smaller groups borrow more information than
larger groups. The global group-specific ESS broadly supports that intuition:
smaller groups such as 2, 6, and 7 borrow substantially more than larger groups
such as 1, 8, 9, and 10. At the same time, group size alone does not determine
borrowing. Group 5 is small but borrows very little on the global scale,
whereas Group 4 is only moderately sized but borrows heavily. Global
conditional ESS therefore reflects group size, within-group variability, and
the group's position in the hierarchical geometry, rather than sample size
alone.

```{r global-ess-plot-scatter}
plot(
  global_group_post_table$sigma_hat2,
  global_group_post_table$ESS_global_post_per_obs,
  pch = 19,
  col = "firebrick",
  xlab = expression(hat(sigma)[i]^2),
  ylab = expression(ESS[theta[i] ~ "|" ~ phi]^glob ~ "(posterior, per observation)"),
  main = expression("Posterior Global Per-Observation ESS vs " * hat(sigma)[i]^2)
)
text(
  global_group_post_table$sigma_hat2,
  global_group_post_table$ESS_global_post_per_obs,
  labels = global_group_post_table$group,
  pos = 3,
  cex = 0.75
)
```

```{r global-ess-plot-sorted}
ord <- order(global_group_post_table$ESS_global_post_per_obs, decreasing = TRUE)
rank_idx <- seq_len(sim$K)

plot(
  rank_idx,
  global_group_post_table$ESS_global_post_per_obs[ord],
  type = "b",
  pch = 19,
  col = "firebrick",
  xaxt = "n",
  xlab = "Group (sorted by posterior global ESS)",
  ylab = expression(ESS[theta[i] ~ "|" ~ phi] ~ "(per observation)"),
  main = "Group-specific ESS: posterior global vs local (posterior ref, per observation)"
)
axis(1, at = rank_idx, labels = global_group_post_table$group[ord])
points(rank_idx, local_group_table$ESS_local_post_ref_per_obs[ord], pch = 17, col = "steelblue")
legend(
  "topright",
  legend = c("Posterior global ESS (per observation)", "Local ESS at posterior reference (per observation)"),
  col = c("firebrick", "steelblue"),
  pch = c(19, 17),
  bty = "n"
)
```

## Sensitivity Analyses

We vary three design dimensions:

- number of groups: \(K \in \{5, 10, 20\}\),
- true between-group variance: \(\tau_0^2 \in \{0.1^2, 0.5^2, 1^2\}\),
- within-group sample size: \(n_i \in \{10, 30, 100\}\).

```{r sensitivity-functions}
fit_and_global_summary <- function(sim_obj, stan_model_hier,
                                   m0, s0, a_tau, b_tau,
                                   iter = 350, warmup = 175,
                                   seed = 1) {
  data_obj <- make_stan_data(sim_obj, m0, s0, a_tau, b_tau)

  fit_obj <- rstan::sampling(
    object = stan_model_hier,
    data = data_obj,
    chains = 1,
    iter = iter,
    warmup = warmup,
    seed = seed,
    refresh = 0,
    control = list(adapt_delta = 0.9)
  )

  theta_draws_obj <- geoess_draws(fit_obj, pars = "theta", as = "matrix")
  mu_draws_obj <- as.numeric(geoess_draws(fit_obj, pars = "mu", as = "matrix"))
  tau2_draws_obj <- as.numeric(geoess_draws(fit_obj, pars = "tau2", as = "matrix"))

  pars_theta_obj <- colnames(theta_draws_obj)
  draws_obj <- cbind(theta_draws_obj, mu = mu_draws_obj, tau2 = tau2_draws_obj)

  ipi_obj <- make_ipi_provider(sim_obj$K, m0, s0, a_tau, b_tau)
  i1_obj <- make_i1_provider(sim_obj$K, sim_obj$sigma_hat2)

  g_obj <- geoess_hier_global(
    Q = "custom",
    draws = draws_obj,
    pars_theta = pars_theta_obj,
    pars_phi = c("mu", "tau2"),
    group_index = seq_len(sim_obj$K),
    Ipi_provider = ipi_obj,
    I1_provider = i1_obj,
    return_by_group = TRUE
  )

  theta_group_per_obs <- g_obj$ess_theta_cond_global
  theta_group_per_group <- theta_group_per_obs / sim_obj$n_i
  data.frame(
    global_ESS_phi_block_per_group = g_obj$ess_phi_global$estimate,
    MCSE_phi_block_per_group = g_obj$ess_phi_global$mcse,
    ess_theta_group_per_obs_mean = mean(theta_group_per_obs),
    ess_theta_group_per_obs_median = median(theta_group_per_obs),
    ess_theta_group_per_obs_q10 = as.numeric(stats::quantile(theta_group_per_obs, probs = 0.10)),
    ess_theta_group_per_obs_q90 = as.numeric(stats::quantile(theta_group_per_obs, probs = 0.90)),
    ess_theta_group_per_group_mean = mean(theta_group_per_group),
    ess_theta_group_per_group_median = median(theta_group_per_group),
    ess_theta_group_per_group_q10 = as.numeric(stats::quantile(theta_group_per_group, probs = 0.10)),
    ess_theta_group_per_group_q90 = as.numeric(stats::quantile(theta_group_per_group, probs = 0.90))
  )
}
```

```{r sensitivity-K}
K_grid <- c(5, 10, 20)

sens_K <- do.call(rbind, lapply(seq_along(K_grid), function(i) {
  K_i <- K_grid[i]
  sim_i <- simulate_hier_data(
    K = K_i,
    n_i = rep(30, K_i),
    mu0 = 0,
    tau0 = 0.5,
    seed = 20261000 + i
  )
  out_i <- fit_and_global_summary(
    sim_obj = sim_i,
    stan_model_hier = stan_model_hier,
    m0 = m0,
    s0 = s0,
    a_tau = a_tau,
    b_tau = b_tau,
    iter = 320,
    warmup = 160,
    seed = 20262000 + i
  )
  cbind(data.frame(K = K_i), out_i)
}))

sens_K
```

```{r sensitivity-tau0}
tau0_grid <- c(0.1, 0.5, 1.0)

sens_tau <- do.call(rbind, lapply(seq_along(tau0_grid), function(i) {
  tau0_i <- tau0_grid[i]
  sim_i <- simulate_hier_data(
    K = 10,
    n_i = rep(30, 10),
    mu0 = 0,
    tau0 = tau0_i,
    seed = 20263000 + i
  )
  out_i <- fit_and_global_summary(
    sim_obj = sim_i,
    stan_model_hier = stan_model_hier,
    m0 = m0,
    s0 = s0,
    a_tau = a_tau,
    b_tau = b_tau,
    iter = 320,
    warmup = 160,
    seed = 20264000 + i
  )
  cbind(data.frame(tau0_sq = tau0_i^2), out_i)
}))

sens_tau
```

```{r sensitivity-ni}
n_i_grid <- c(10, 30, 100)

sens_n <- do.call(rbind, lapply(seq_along(n_i_grid), function(i) {
  n_i_val <- n_i_grid[i]
  sim_i <- simulate_hier_data(
    K = 10,
    n_i = rep(n_i_val, 10),
    mu0 = 0,
    tau0 = 0.5,
    seed = 20265000 + i
  )
  out_i <- fit_and_global_summary(
    sim_obj = sim_i,
    stan_model_hier = stan_model_hier,
    m0 = m0,
    s0 = s0,
    a_tau = a_tau,
    b_tau = b_tau,
    iter = 320,
    warmup = 160,
    seed = 20266000 + i
  )
  cbind(data.frame(n_i = n_i_val), out_i)
}))

sens_n
```

```{r sensitivity-plot}
plot(
  sens_tau$tau0_sq,
  sens_tau$global_ESS_phi_block_per_group,
  type = "b",
  pch = 19,
  col = "firebrick",
  xlab = expression(tau[0]^2),
  ylab = "Posterior global ESS for phi (per group)",
  main = expression("Global Per-Group Block ESS vs " * tau[0]^2)
)
```

```{r sensitivity-plot-theta}
plot(
  sens_tau$tau0_sq,
  sens_tau$ess_theta_group_per_obs_median,
  type = "b",
  pch = 19,
  col = "steelblue",
  xlab = expression(tau[0]^2),
  ylab = expression("Median across groups of " * ESS[theta[i] ~ "|" ~ phi]^glob ~ "(per observation)"),
  main = expression("Distribution of Group Per-Observation ESS vs " * tau[0]^2)
)
segments(
  x0 = sens_tau$tau0_sq,
  y0 = sens_tau$ess_theta_group_per_obs_q10,
  x1 = sens_tau$tau0_sq,
  y1 = sens_tau$ess_theta_group_per_obs_q90,
  col = "gray40",
  lwd = 2
)
```

## Practical Interpretation

- `ESS_phi` is a block ESS for the hyperparameters in units of groups, so it
  summarizes the strength of the shared borrowing structure itself.
- `ESS_{theta_i|phi}^{(\mathrm{obs})}` measures how much borrowing group \(i\)
  receives relative to one observation.
- `ESS_{theta_i|phi}^{(\mathrm{grp})}` measures the same borrowing relative to
  the full data block of group \(i\), which makes it the more natural scale for
  comparing groups with different sample sizes.
- The contrast between Groups 2 and 5 is a useful diagnostic: a very small
  local ESS can coexist with a large global ESS, and vice versa, when the
  chosen reference point is not representative of the broader hierarchical
  geometry.
- For reporting, the global conditional ESS is usually the more stable summary,
  while the local conditional ESS is most useful as a pointwise diagnostic.

At a high level, this is conceptually aligned with Morita et al. (2012): ESS in
hierarchical models is level-specific and design-dependent. The geometric
framework here emphasizes curvature-based local definitions plus explicit global
averaging under a stated \(Q\).

## Discussion

This vignette shows end-to-end hierarchical ESS for information borrowing with:

- local block ESS for \(\phi=(\mu,\tau^2)\) in units of groups,
- local conditional ESS for each \(\theta_i\mid\phi\) in units of
  observations and in units of groups,
- global ESS under prior and posterior averaging on both scales.

Using plug-in \(\widehat{\sigma}_i^2\) keeps the workflow simple while still
capturing the main geometric features of hierarchical borrowing. The examples
above show that borrowing strength is not a single universal number: it depends
on the model level being summarized, the unit of comparison, the reference
point, and the weighting distribution. The sensitivity analyses over \(K\),
\(\tau_0^2\), and \(n_i\) reinforce the same message: hyperparameter ESS is
relatively stable, whereas group-level conditional ESS can change materially
with hierarchical design and heterogeneity.
