---
title: "Logistic regression with Gaussian prior"
output: html_vignette
vignette: >
  %\VignetteIndexEntry{Logistic regression with Gaussian prior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
set.seed(20260218)

library(geoess)
library(rstan)

rstan::rstan_options(auto_write = FALSE)
options(mc.cores = 2)
```

## Motivation

Logistic regression with Gaussian prior is a canonical non-conjugate setting,
so the effective sample size (ESS) is not available from closed-form posterior
updating. This vignette shows how to compute geometric ESS using `geoess` by
combining analytic prior curvature with numerically evaluated per-observation
Fisher information.

## Model and Priors

We consider logistic regression with two Gaussian joint priors for
\(\beta = (\beta_0, \beta_1)\): an independent prior
\(\beta \sim \mathcal{N}(0, \tau^2 I_d)\) and a correlated prior
\(\beta \sim \mathcal{N}(0, \Sigma_\rho)\), where
\[
\Sigma_\rho = \tau^2
\begin{pmatrix}
1 & \rho \\
\rho & 1
\end{pmatrix}.
\]
Throughout the main analysis we use \(\tau = 1\) and, for the correlated
scenario, \(\rho = 0.6\).

The sampling model is

\[
y_i \mid x_i,\beta \sim \text{Bernoulli}\!\left(\operatorname{logit}^{-1}(x_i^\top\beta)\right),
\qquad i=1,\dots,n,
\]
\[
I_\pi(\beta) = -\nabla^2 \log \pi(\beta).
\]

For the independent Gaussian prior,
\[
I_\pi(\beta) = \tau^{-2} I_d,
\]
and for the correlated prior,
\[
I_{\pi,\rho}(\beta) = \Sigma_\rho^{-1}.
\]

For logistic regression, the per-observation Fisher information at a reference
point \(\beta^*\) is estimated by
\[
I_1(\beta^*) \approx \frac{1}{n}\sum_{i=1}^n w_i x_i x_i^\top,
\qquad w_i = p_i(1-p_i),\quad p_i = \operatorname{logit}^{-1}(x_i^\top\beta^*).
\]

Local ESS (trace form) is
\[
\mathrm{ESS}_{\text{local}}(\beta^*) = \frac{1}{d}\operatorname{tr}\!\left(I_1(\beta^*)^{-1}I_\pi\right).
\]

## Data Generating Setup

```{r data-generating-setup}
set.seed(20260218)

n <- 200
# d = 2 by default: intercept + one centered standardized covariate.
# The centering ensures that at the symmetry reference beta = (0, 0), the
# per-observation Fisher information is axis-aligned, so an independent prior
# yields coordinate eigenvectors while a correlated prior rotates them.
d <- 2
beta_true <- c(-0.5, 1.0)
tau <- 1

x_raw <- rnorm(n)
x_std <- as.numeric(scale(x_raw))
X <- cbind(`(Intercept)` = 1, x = x_std)

eta_true <- as.vector(X %*% beta_true)
p_true <- plogis(eta_true)
y <- rbinom(n, size = 1, prob = p_true)

# Reusable helpers for this vignette
fisher_logistic_per_obs <- function(beta, X) {
  eta <- as.vector(X %*% beta)
  p <- plogis(eta)
  w <- p * (1 - p)
  Xw <- sweep(X, 1, w, "*")
  I1 <- crossprod(X, Xw) / nrow(X)
  0.5 * (I1 + t(I1))
}

prior_covariance_gaussian <- function(d, tau, rho = 0) {
  if (d == 2) {
    return(tau^2 * matrix(c(1, rho, rho, 1), nrow = 2, byrow = TRUE))
  }
  if (abs(rho) > sqrt(.Machine$double.eps)) {
    stop("Correlated Gaussian prior is currently implemented only for d = 2.")
  }
  diag(tau^2, d)
}

prior_curvature_gaussian <- function(d, tau, rho = 0) {
  solve(prior_covariance_gaussian(d = d, tau = tau, rho = rho))
}

draw_prior_beta <- function(n_draws, d, tau = NULL, rho = 0, Sigma_prior = NULL) {
  if (is.null(Sigma_prior)) {
    if (is.null(tau)) stop("Provide either tau or Sigma_prior.")
    Sigma_prior <- prior_covariance_gaussian(d = d, tau = tau, rho = rho)
  }
  R <- chol(Sigma_prior)
  out <- matrix(rnorm(n_draws * d), ncol = d) %*% R
  colnames(out) <- colnames(X)
  out
}

data.frame(
  n = n,
  d = d,
  beta_true_intercept = beta_true[1],
  beta_true_slope = beta_true[2],
  event_rate = mean(y)
)
```

This simulation uses \(n=200\) observations with \(d=2\) predictors: an
intercept with coefficient \(\beta_0\) and one standardized covariate with
coefficient \(\beta_1\). The true coefficient vector is
\(\beta_{\mathrm{true}} = (-0.5, 1.0)\), and the realized event rate is
`r format(mean(y), digits = 3)`.

## Posterior Sampling via Stan

```{r stan-model-and-fit}
stan_code <- "
data {
  int<lower=1> N;
  int<lower=1> D;
  matrix[N, D] X;
  array[N] int<lower=0, upper=1> y;
  real<lower=0> tau;
}
parameters {
  vector[D] beta;
}
model {
  beta ~ normal(0, tau);
  y ~ bernoulli_logit(X * beta);
}
generated quantities {
  vector[N] log_lik;
  for (i in 1:N) {
    log_lik[i] = bernoulli_logit_lpmf(y[i] | dot_product(row(X, i), beta));
  }
}
"

stan_model_logit <- rstan::stan_model(model_code = stan_code)

stan_data <- list(N = n, D = d, X = X, y = y, tau = tau)
fit_post <- rstan::sampling(
  object = stan_model_logit,
  data = stan_data,
  chains = 2,
  iter = 800,
  warmup = 400,
  seed = 20260218,
  refresh = 0,
  control = list(adapt_delta = 0.9)
)

beta_draws <- geoess_draws(fit_post, pars = "beta", as = "matrix")
colnames(beta_draws) <- colnames(X)

# Posterior mean as a local reference point (via geoess helper)
beta_post_mean <- as.numeric(geoess_theta_ref(beta_draws, ref = "posterior_mean"))
names(beta_post_mean) <- colnames(X)

# Prior mean reference point
beta_prior_mean <- rep(0, d)
names(beta_prior_mean) <- colnames(X)

rbind(
  prior_mean = beta_prior_mean,
  posterior_mean = beta_post_mean
)
```

```{r stan-model-and-fit-correlated}
rho <- 0.6
Sigma_prior_rho <- prior_covariance_gaussian(d = d, tau = tau, rho = rho)

stan_code_corr <- "
data {
  int<lower=1> N;
  int<lower=1> D;
  matrix[N, D] X;
  array[N] int<lower=0, upper=1> y;
  cov_matrix[D] Sigma_prior;
}
transformed data {
  vector[D] beta_mean = rep_vector(0, D);
}
parameters {
  vector[D] beta;
}
model {
  beta ~ multi_normal(beta_mean, Sigma_prior);
  y ~ bernoulli_logit(X * beta);
}
generated quantities {
  vector[N] log_lik;
  for (i in 1:N) {
    log_lik[i] = bernoulli_logit_lpmf(y[i] | dot_product(row(X, i), beta));
  }
}
"

stan_model_logit_corr <- rstan::stan_model(model_code = stan_code_corr)
stan_data_corr <- list(N = n, D = d, X = X, y = y, Sigma_prior = Sigma_prior_rho)

fit_post_rho <- rstan::sampling(
  object = stan_model_logit_corr,
  data = stan_data_corr,
  chains = 2,
  iter = 800,
  warmup = 400,
  seed = 20260220,
  refresh = 0,
  control = list(adapt_delta = 0.9)
)

beta_draws_rho <- geoess_draws(fit_post_rho, pars = "beta", as = "matrix")
colnames(beta_draws_rho) <- colnames(X)

beta_post_mean_rho <- as.numeric(geoess_theta_ref(beta_draws_rho, ref = "posterior_mean"))
names(beta_post_mean_rho) <- colnames(X)

rbind(
  independent_posterior_mean = beta_post_mean,
  correlated_posterior_mean = beta_post_mean_rho
)
```

For more accurate posterior summaries and global ESS estimates, increase
`iter`, `warmup`, or the number of chains.

## Local ESS

```{r local-ess}
Ipi <- prior_curvature_gaussian(d = d, tau = tau)

I1_at_prior_mean <- fisher_logistic_per_obs(beta_prior_mean, X)
I1_at_post_mean <- fisher_logistic_per_obs(beta_post_mean, X)

# Local ESS computed via geoess conventions: (1/d) tr(I1^{-1} Ipi)
local_ess_prior_mean <- geoess_trace(I1_at_prior_mean, Ipi)
local_ess_post_mean <- geoess_trace(I1_at_post_mean, Ipi)

local_ess_table <- data.frame(
  reference_point = c("Prior mean (0, 0)", "Posterior mean"),
  local_ess = c(local_ess_prior_mean, local_ess_post_mean)
)
invisible(NULL)
```

```{r ess-spectrum}
# ESS spectrum at the symmetry reference beta = (0, 0).
# Because the covariate is centered, I1 is diagonal here and the independent
# Gaussian prior yields coordinate eigenvectors.
I1_at_symmetry_ref <- fisher_logistic_per_obs(beta_prior_mean, X)
spec_post_mean <- geoess_spectrum(I1_at_symmetry_ref, Ipi)

spectrum_values <- data.frame(
  direction = paste0("eig", seq_along(spec_post_mean$values)),
  eigenvalue = spec_post_mean$values
)

eigenvectors <- as.data.frame(spec_post_mean$vectors)
colnames(eigenvectors) <- paste0("eig", seq_len(ncol(eigenvectors)))
param_names <- rownames(spec_post_mean$vectors)
if (is.null(param_names)) {
  param_names <- colnames(X)
}
eigenvectors$parameter <- param_names
eigenvectors <- eigenvectors[, c("parameter", setdiff(names(eigenvectors), "parameter"))]

invisible(NULL)
```

## Correlated Gaussian Prior

We now consider a correlated Gaussian prior with
\[
\beta \sim \mathcal{N}(0, \Sigma_\rho),
\qquad
\Sigma_\rho = \tau^2
\begin{pmatrix}
1 & \rho \\
\rho & 1
\end{pmatrix},
\]
using \(\rho = 0.6\). We keep the same dataset, but for the spectral
comparison we evaluate ESS at the same symmetry reference point
\(\beta = (0, 0)\). This makes the contrast with the independent prior clear:
the likelihood information remains axis-aligned, and the anisotropy arises from
the prior correlation.

```{r correlated-prior}
Ipi_rho <- prior_curvature_gaussian(d = d, tau = tau, rho = rho)
I1_at_symmetry_ref_rho <- fisher_logistic_per_obs(beta_prior_mean, X)

e_beta0 <- c(1, 0)
e_beta1 <- c(0, 1)

# Per-coordinate ESS summaries under the correlated prior.
ess_beta0_rho <- geoess_dir(I1_at_symmetry_ref_rho, Ipi_rho, e_beta0)
ess_beta1_rho <- geoess_dir(I1_at_symmetry_ref_rho, Ipi_rho, e_beta1)

coord_ess_rho <- data.frame(
  parameter = c("beta0", "beta1"),
  marginal_ess = c(ess_beta0_rho, ess_beta1_rho)
)

spec_rho <- geoess_spectrum(I1_at_symmetry_ref_rho, Ipi_rho)

eigenvalues_rho <- data.frame(
  direction = paste0("eig", seq_along(spec_rho$values)),
  eigenvalue = spec_rho$values
)

eigenvectors_rho <- as.data.frame(spec_rho$vectors)
colnames(eigenvectors_rho) <- paste0("eig", seq_len(ncol(eigenvectors_rho)))
param_names_rho <- rownames(spec_rho$vectors)
if (is.null(param_names_rho)) {
  param_names_rho <- colnames(X)
}
eigenvectors_rho$parameter <- param_names_rho
eigenvectors_rho <- eigenvectors_rho[, c("parameter", setdiff(names(eigenvectors_rho), "parameter"))]

invisible(NULL)
```

## Global ESS

We report two weighting distributions:

- Prior-averaged global ESS: \(Q = \pi(\beta)\).
- Posterior-averaged global ESS: \(Q = \pi(\beta \mid y)\).
- We report both the independent prior (\(\rho = 0\)) and the correlated prior
  (\(\rho = 0.6\)).

```{r global-ess}
set.seed(20260219)

n_mc <- min(400, nrow(beta_draws))
idx_post <- sample(seq_len(nrow(beta_draws)), size = n_mc)
beta_post_mc <- beta_draws[idx_post, , drop = FALSE]
beta_prior_mc <- draw_prior_beta(n_draws = n_mc, d = d, tau = tau, rho = 0)

set.seed(20260220)
idx_post_rho <- sample(seq_len(nrow(beta_draws_rho)), size = n_mc)
beta_post_mc_rho <- beta_draws_rho[idx_post_rho, , drop = FALSE]
beta_prior_mc_rho <- draw_prior_beta(
  n_draws = n_mc,
  d = d,
  Sigma_prior = Sigma_prior_rho
)

global_post <- geoess_global(
  Q = list(draws = beta_post_mc, type = "posterior"),
  I1 = function(theta) fisher_logistic_per_obs(theta, X),
  Ipi = prior_curvature_gaussian(d = d, tau = tau, rho = 0),
  method = "trace"
)

global_prior <- geoess_global(
  Q = list(draws = beta_prior_mc, type = "prior"),
  I1 = function(theta) fisher_logistic_per_obs(theta, X),
  Ipi = prior_curvature_gaussian(d = d, tau = tau, rho = 0),
  method = "trace"
)

global_post_rho <- geoess_global(
  Q = list(draws = beta_post_mc_rho, type = "posterior"),
  I1 = function(theta) fisher_logistic_per_obs(theta, X),
  Ipi = Ipi_rho,
  method = "trace"
)

global_prior_rho <- geoess_global(
  Q = list(draws = beta_prior_mc_rho, type = "prior"),
  I1 = function(theta) fisher_logistic_per_obs(theta, X),
  Ipi = Ipi_rho,
  method = "trace"
)

global_ess_table <- data.frame(
  prior_structure = c(
    "Independent (rho = 0)",
    "Independent (rho = 0)",
    "Correlated (rho = 0.6)",
    "Correlated (rho = 0.6)"
  ),
  weighting_distribution = c(
    "Prior (Q = pi(beta))",
    "Posterior (Q = pi(beta | y))",
    "Prior (Q = pi(beta))",
    "Posterior (Q = pi(beta | y))"
  ),
  ess_global = c(
    global_prior$estimate,
    global_post$estimate,
    global_prior_rho$estimate,
    global_post_rho$estimate
  ),
  mcse = c(
    global_prior$mcse,
    global_post$mcse,
    global_prior_rho$mcse,
    global_post_rho$mcse
  ),
  n_draws = c(
    global_prior$n,
    global_post$n,
    global_prior_rho$n,
    global_post_rho$n
  )
)
invisible(NULL)
```

## ESS Summary

```{r ess-summary}
I1_at_prior_mean_rho <- fisher_logistic_per_obs(beta_prior_mean, X)
I1_at_post_mean_rho <- fisher_logistic_per_obs(beta_post_mean_rho, X)

local_ess_prior_mean_rho <- geoess_trace(I1_at_prior_mean_rho, Ipi_rho)
local_ess_post_mean_rho <- geoess_trace(I1_at_post_mean_rho, Ipi_rho)

format_vec <- function(v, digits = 3) {
  paste0("(", paste(format(round(v, digits), nsmall = digits, trim = TRUE), collapse = ", "), ")")
}

panel_a_display <- data.frame(
  Scenario = c("Independent", sprintf("Correlated (rho = %.1f)", rho)),
  `Local ESS<br>Prior mean` = c(local_ess_prior_mean, local_ess_prior_mean_rho),
  `Local ESS<br>Posterior mean` = c(local_ess_post_mean, local_ess_post_mean_rho),
  `Global ESS<br>Prior-avg` = c(global_prior$estimate, global_prior_rho$estimate),
  `Global ESS<br>Posterior-avg` = c(global_post$estimate, global_post_rho$estimate),
  check.names = FALSE
)
panel_a_display[-1] <- lapply(panel_a_display[-1], round, 1)

panel_b_display <- data.frame(
  Scenario = c("Independent", sprintf("Correlated (rho = %.1f)", rho)),
  `Marginal ESS<br>beta0` = c(
    geoess_dir(I1_at_symmetry_ref, Ipi, c(1, 0)),
    ess_beta0_rho
  ),
  `Marginal ESS<br>beta1` = c(
    geoess_dir(I1_at_symmetry_ref, Ipi, c(0, 1)),
    ess_beta1_rho
  ),
  `Direction 1<br>(eigenvector)` = c(
    "$(1,0)$",
    paste0("$", format_vec(spec_rho$vectors[, 1]), "$")
  ),
  `Directional ESS 1<br>(eigenvalue)` = c(spec_post_mean$values[1], spec_rho$values[1]),
  `Direction 2<br>(eigenvector)` = c(
    "$(0,1)$",
    paste0("$", format_vec(spec_rho$vectors[, 2]), "$")
  ),
  `Directional ESS 2<br>(eigenvalue)` = c(spec_post_mean$values[2], spec_rho$values[2]),
  check.names = FALSE
)
panel_b_display[c(2, 3, 5, 7)] <- lapply(panel_b_display[c(2, 3, 5, 7)], round, 1)

knitr::kable(
  panel_a_display,
  escape = FALSE,
  caption = paste(
    "Panel A: local and global ESS of the joint prior for logistic regression.",
    "Local ESS is evaluated at the prior and posterior means; global ESS is",
    "reported under prior and posterior averaging."
  )
)

knitr::kable(
  panel_b_display,
  escape = FALSE,
  caption = paste(
    "Panel B: local marginal and directional ESS at the prior mean.",
    "For the independent prior, the geometry is coordinate-aligned, so the",
    "marginal and directional ESS coincide. Under the correlated prior, the",
    "eigenvectors rotate away from the coordinate axes and the directional ESS",
    "reveals stronger anisotropy."
  )
)
```

Panel A summarizes the local and global ESS for the joint prior under both
independent and correlated Gaussian scenarios. A useful pattern appears in both
cases: local ESS is larger at the posterior mean than at the prior mean, but
global ESS is larger under prior averaging than under posterior averaging. This
is not a contradiction. ESS measures prior informativeness relative to the unit
of data, not absolute informativeness of a point. At the prior mean, the data
are locally more informative than at the posterior mean; equivalently, the
prior has more local influence at the posterior mean. For the global summaries,
the prior distribution is broader than the posterior and therefore places more
weight on tail regions where the logistic likelihood is flatter. Those
low-information tail regions inflate the prior-averaged ESS relative to the
posterior-averaged ESS.

Panel B reports local marginal and directional ESS at the prior mean. Under the
independent prior, the centered design makes the relative information geometry
essentially coordinate-aligned, so the marginal ESS and the directional ESS are
the same. Under the correlated prior, the principal directions rotate away from
the coordinate axes and the directional ESS becomes notably anisotropic. In
that setting, the marginal ESS is similar across \(\beta_0\) and \(\beta_1\),
but the directional ESS differs sharply across the two eigen-directions. This
illustrates why multiple ESS summaries are useful even in a two-parameter
example.

## Sensitivity Analyses

We vary \(\tau \in \{0.5, 1, 2\}\), keeping the same dataset, and compare
independent and correlated Gaussian priors. Each setting runs a short Stan fit
to keep vignette runtime moderate.

```{r sensitivity-tau}
run_tau_sensitivity <- function(tau_value, rho_value, stan_model_indep,
                                stan_model_corr, X, y, seed_offset = 0) {
  d_loc <- ncol(X)
  Sigma_tau <- prior_covariance_gaussian(d = d_loc, tau = tau_value, rho = rho_value)

  if (abs(rho_value) < sqrt(.Machine$double.eps)) {
    data_tau <- list(N = nrow(X), D = d_loc, X = X, y = y, tau = tau_value)
    fit_tau <- rstan::sampling(
      object = stan_model_indep,
      data = data_tau,
      chains = 1,
      iter = 500,
      warmup = 250,
      seed = 20261000 + seed_offset,
      refresh = 0,
      control = list(adapt_delta = 0.9)
    )
  } else {
    data_tau <- list(N = nrow(X), D = d_loc, X = X, y = y, Sigma_prior = Sigma_tau)
    fit_tau <- rstan::sampling(
      object = stan_model_corr,
      data = data_tau,
      chains = 1,
      iter = 500,
      warmup = 250,
      seed = 20261000 + seed_offset,
      refresh = 0,
      control = list(adapt_delta = 0.9)
    )
  }

  draws_tau <- geoess_draws(fit_tau, pars = "beta", as = "matrix")
  colnames(draws_tau) <- colnames(X)

  beta_mean_tau <- as.numeric(geoess_theta_ref(draws_tau, ref = "posterior_mean"))
  Ipi_tau <- prior_curvature_gaussian(d = d_loc, tau = tau_value, rho = rho_value)
  local_tau <- geoess_trace(fisher_logistic_per_obs(beta_mean_tau, X), Ipi_tau)

  m_tau <- min(300, nrow(draws_tau))
  set.seed(20262000 + 10 * seed_offset)
  idx_tau <- sample(seq_len(nrow(draws_tau)), size = m_tau)
  draws_post_tau <- draws_tau[idx_tau, , drop = FALSE]
  set.seed(20263000 + 10 * seed_offset)
  draws_prior_tau <- draw_prior_beta(
    n_draws = m_tau,
    d = d_loc,
    Sigma_prior = Sigma_tau
  )

  g_post_tau <- geoess_global(
    Q = list(draws = draws_post_tau, type = "posterior"),
    I1 = function(theta) fisher_logistic_per_obs(theta, X),
    Ipi = Ipi_tau,
    method = "trace"
  )

  g_prior_tau <- geoess_global(
    Q = list(draws = draws_prior_tau, type = "prior"),
    I1 = function(theta) fisher_logistic_per_obs(theta, X),
    Ipi = Ipi_tau,
    method = "trace"
  )

  data.frame(
    prior_structure = if (abs(rho_value) < sqrt(.Machine$double.eps)) {
      "Independent (rho = 0)"
    } else {
      sprintf("Correlated (rho = %.1f)", rho_value)
    },
    rho = rho_value,
    tau = tau_value,
    local_ess_at_post_mean = local_tau,
    global_ess_prior = g_prior_tau$estimate,
    mcse_prior = g_prior_tau$mcse,
    global_ess_posterior = g_post_tau$estimate,
    mcse_posterior = g_post_tau$mcse
  )
}

tau_grid <- c(0.5, 1, 2)
rho_grid <- c(0, 0.6)
sensitivity_tau <- do.call(rbind, lapply(seq_along(rho_grid), function(j) {
  do.call(
    rbind,
    lapply(seq_along(tau_grid), function(i) {
      run_tau_sensitivity(
        tau_value = tau_grid[i],
        rho_value = rho_grid[j],
        stan_model_indep = stan_model_logit,
        stan_model_corr = stan_model_logit_corr,
        X = X,
        y = y,
        seed_offset = 10 * j + i
      )
    })
  )
}))

sensitivity_tau
```

Optional extension to higher dimension (e.g., \(d=5\)):

```{r optional-d5, eval=FALSE}
# Example sketch for d = 5 (intercept + 4 covariates)
# x_mat <- scale(matrix(rnorm(n * 4), nrow = n, ncol = 4))
# X5 <- cbind(1, x_mat)
# beta_true5 <- c(-0.5, 1.0, -0.75, 0.5, 0.25)
# y5 <- rbinom(n, 1, plogis(X5 %*% beta_true5))
# Then reuse the same Stan model with D = 5 and repeat local/global ESS steps.
```

## Discussion

This non-conjugate example illustrates the geometric ESS workflow in a model
where curvature depends on both the reference point and the weighting
distribution. The independent and correlated priors show the same qualitative
local/global reversal, but they differ substantially in their directional
structure. The independent prior is essentially coordinate-aligned, whereas the
correlated prior rotates the principal directions and induces stronger
anisotropy. Together, these summaries show that a single scalar ESS is often
not enough to describe prior informativeness fully in nonconjugate regression
models.
