---
title: "Numerical priors"
output: html_vignette
vignette: >
  %\VignetteIndexEntry{Numerical priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
set.seed(123)

library(geoess)
```

## Theory recap

Local ESS at a reference point $\theta^*$:

$$
ESS(\theta^*) = \frac{1}{d}\,\mathrm{tr}\{ I_1(\theta^*)^{-1} I_\pi(\theta^*) \}
$$

For mixture priors, geoess now also supports \emph{component-local} ESS
reporting at component-specific reference points $\theta_k^*$:
$$
ESS_k(\theta_k^*) = \frac{1}{d}\,\mathrm{tr}\{ I_1(\theta_k^*)^{-1} I_{\pi_k}(\theta_k^*) \},
$$
with optional attribution scaling:
$$
ESS^{(w)}_k = w_k\,ESS_k.
$$
Here, $w_k$ is the mixture weight.
If $\theta^*$ is a tied saddle point of components $k$ and $k+1$, the local ESS
is defined as the average of the tied components' weighted ESS values:
\begin{equation}
  ESS(\theta^*)=[w_{k} \cdot ESS_{k}(\theta^*)+w_{k+1} \cdot ESS_{k+1}(\theta^*)]/2 \tag{1}.
\end{equation}
This convention bypasses the saddle's negative joint curvature, ensuring a
non-negative and practically interpretable ESS at transition boundaries.

Global ESS under $Q$:

$$
ESS_Q = \mathbb{E}_{\theta \sim Q}[ ESS(\theta) ].
$$

Here the prior is **unknown analytically** and only available via **samples**.
We therefore:

1. Draw $\theta$ from the prior and fit a parametric mixture $q(\theta)$.
2. Use component curvature together with weighted hard assignment to compute
   local ESS for mixture priors.
3. Approximate global ESS by averaging those local ESS values under $Q$ using
   the same draws.
4. Use likelihood reweighting of prior draws to approximate posterior quantities.

## Motivation

To evaluate ESS under complex prior structures, we use a bimodal prior and
compare three types of summaries:

- reference-point-specific local ESS,
- distribution-averaged global ESS, and
- existing ESS estimates from `RBesT`.

The goal is to show that for a complex prior, no single ESS summary is
adequate on its own.

## Simulation Design

We simulate the prior as a two-component mixture on the real line:

$$
\pi(\theta) = 0.3\,\mathcal{N}(\mu_1, s_1^2) + 0.7\,\text{Beta}(a,b)\,\mathbf{1}_{(0,1)}(\theta).
$$

The Beta component contributes only on $(0,1)$, so the mixture has two local modes.
We do **not** assume this mixture is known; we only observe samples from it.
The parameters are chosen to make the two modes well-separated so local ESS
changes with the reference point.

```{r prior-simulation}
# Mixture prior parameters
w <- c(0.3, 0.7)
# Separated components to make reference-point effects visible.
mu1 <- 0.1
s1 <- 0.08
a <- 10
b <- 2

n_draws <- 1000
set.seed(123)
z <- rbinom(n_draws, size = 1, prob = w[1])
prior_draws <- ifelse(
  z == 1,
  rnorm(n_draws, mean = mu1, sd = s1),
  rbeta(n_draws, a, b)
)

summary(prior_draws)
```

## Helper utilities

```{r helpers}
logsumexp <- function(x) {
  m <- max(x)
  m + log(sum(exp(x - m)))
}

find_modes <- function(x, k = 2, bw = "nrd0") {
  d <- density(x, bw = bw)
  y <- d$y
  idx <- which(y > c(-Inf, y[-length(y)]) & y > c(y[-1], -Inf))
  if (length(idx) == 0L) return(rep(NA_real_, k))
  ord <- idx[order(y[idx], decreasing = TRUE)]
  modes <- d$x[ord][seq_len(min(k, length(ord)))]
  if (length(modes) < k) modes <- c(modes, rep(NA_real_, k - length(modes)))
  modes
}

find_saddle_between_modes <- function(x, bw = "nrd0") {
  d <- density(x, bw = bw)
  y <- d$y
  idx_max <- which(y > c(-Inf, y[-length(y)]) & y > c(y[-1], -Inf))
  if (length(idx_max) < 2L) return(NA_real_)

  top2 <- idx_max[order(y[idx_max], decreasing = TRUE)[1:2]]
  lo <- min(top2)
  hi <- max(top2)
  if ((hi - lo) < 2L) return(NA_real_)

  idx_min <- which(y < c(Inf, y[-length(y)]) & y < c(y[-1], Inf))
  idx_between <- idx_min[idx_min > lo & idx_min < hi]
  j <- if (length(idx_between) > 0L) {
    idx_between[which.min(y[idx_between])]
  } else {
    lo + which.min(y[lo:hi]) - 1L
  }
  d$x[j]
}

find_mix_resp_tie <- function(mix, comp = NULL, tol = 1e-10) {
  if (is.null(mix$weights) || is.null(mix$mu) || is.null(mix$Sigma)) {
    return(NA_real_)
  }
  w <- as.numeric(mix$weights)
  w <- w / sum(w)
  K <- length(w)
  if (K < 2L) return(NA_real_)

  if (is.null(comp)) {
    comp <- order(w, decreasing = TRUE)[1:2]
  }
  a <- comp[1]
  b <- comp[2]

  mu_a <- as.numeric(mix$mu[[a]])[1]
  mu_b <- as.numeric(mix$mu[[b]])[1]
  sd_a <- sqrt(as.numeric(mix$Sigma[[a]][1, 1]))
  sd_b <- sqrt(as.numeric(mix$Sigma[[b]][1, 1]))

  f <- function(x) {
    (log(w[a]) + dnorm(x, mean = mu_a, sd = sd_a, log = TRUE)) -
      (log(w[b]) + dnorm(x, mean = mu_b, sd = sd_b, log = TRUE))
  }

  lo <- min(mu_a, mu_b)
  hi <- max(mu_a, mu_b)
  flo <- f(lo)
  fhi <- f(hi)

  if (abs(flo) <= tol) return(lo)
  if (abs(fhi) <= tol) return(hi)

  if (!is.finite(flo) || !is.finite(fhi) || flo * fhi > 0) {
    lo <- min(mu_a - 6 * sd_a, mu_b - 6 * sd_b)
    hi <- max(mu_a + 6 * sd_a, mu_b + 6 * sd_b)
    flo <- f(lo)
    fhi <- f(hi)
    if (!is.finite(flo) || !is.finite(fhi) || flo * fhi > 0) {
      return(NA_real_)
    }
  }

  stats::uniroot(f, lower = lo, upper = hi, tol = tol)$root
}

weighted_mode <- function(x, w, bw = "nrd0") {
  w <- w / sum(w)
  d <- density(x, weights = w, bw = bw)
  d$x[which.max(d$y)]
}

rbest_mixfit_best <- function(draws, type, Nc_grid = 1:2, aic_k = 2) {
  if (!requireNamespace("RBesT", quietly = TRUE)) return(NULL)
  if (!isTRUE(exists("mixfit", envir = asNamespace("RBesT"), mode = "function"))) {
    return(NULL)
  }
  draws <- as.numeric(draws)
  draws <- draws[is.finite(draws)]
  if (length(draws) < 5L) return(NULL)
  f <- get("mixfit", envir = asNamespace("RBesT"))
  fits <- lapply(Nc_grid, function(nc) {
    args <- list(sample = draws, type = type, Nc = nc, verbose = FALSE)
    tryCatch(do.call(f, args), error = function(e) NULL)
  })
  aic <- vapply(
    fits,
    function(fit) if (is.null(fit)) Inf else AIC(fit, k = aic_k),
    numeric(1)
  )
  if (!any(is.finite(aic))) return(NULL)
  fits[[which.min(aic)]]
}

rbest_ess_from_draws <- function(draws, sigma, Nc_grid = 1:2) {
  if (!requireNamespace("RBesT", quietly = TRUE)) {
    return(c(elir = NA_real_, moment = NA_real_, morita = NA_real_))
  }
  mix <- rbest_mixfit_best(draws, type = "norm", Nc_grid = Nc_grid)
  if (is.null(mix)) return(c(elir = NA_real_, moment = NA_real_, morita = NA_real_))
  vapply(
    c("elir", "moment", "morita"),
    function(m) tryCatch(RBesT::ess(mix, method = m, sigma = sigma),
                         error = function(e) NA_real_),
    numeric(1)
  )
}
```

## Sample-based prior approximation

We approximate the unknown prior by a **normal mixture** fit to the samples.
This is a generic choice that can represent multi-modality even when the true
prior is not normal. We **fix \(K=2\)** and select the fit by BIC within that
two-component family so the approximation preserves the bimodality and the
local ESS varies across reference points.

```{r prior-fit}
I1 <- matrix(1, nrow = 1, ncol = 1)  # placeholder, replaced below
mix <- geoess_automixfit(matrix(prior_draws, ncol = 1), family = "norm",
                         K_grid = 2, criterion = "BIC")
mix$K_selected
```

## Likelihood model

We use a Normal likelihood with known $\sigma$:

$$
Y_i \mid \theta \sim \mathcal{N}(\theta, \sigma^2), \quad I_1 = 1/\sigma^2.
$$

```{r likelihood}
N <- 10
sigma <- 1
I1 <- matrix(1 / sigma^2, nrow = 1, ncol = 1)

# Generate data from a value near the normal component mode
theta_true <- 0.2
set.seed(123)
y <- rnorm(N, mean = theta_true, sd = sigma)
y
loglik_draw <- function(theta) {
  sum(dnorm(y, mean = theta, sd = sigma, log = TRUE))
}
```

## Local ESS: effect of reference point

We compare local ESS at different reference points:

- prior sample mean
- two local prior modes
- a misspecified point at the saddle between the two dominant modes
- posterior mean (importance-weighted)
- posterior mode (importance-weighted)

```{r local-ess}
# Prior reference points
prior_mean <- mean(prior_draws)
prior_modes <- find_modes(prior_draws, k = 2)
prior_misspec <- find_mix_resp_tie(mix)
if (!is.finite(prior_misspec)) {
  prior_misspec <- find_saddle_between_modes(prior_draws)
}
if (!is.finite(prior_misspec)) {
  prior_misspec <- mean(prior_modes, na.rm = TRUE)
}

# Approximate posterior weights via importance sampling
logw <- vapply(prior_draws, loglik_draw, numeric(1))
post_w <- exp(logw - logsumexp(logw))

post_mean <- sum(post_w * prior_draws)
post_mode <- weighted_mode(prior_draws, post_w)

# Local ESS at reference points (using sample-based mixture for Ipi)
ref_points <- c(prior_mean, prior_modes[1], prior_modes[2], prior_misspec,
                post_mean, post_mode)
ref_labels <- c("prior_mean", "prior_mode_1", "prior_mode_2",
                "prior_misspecified", "posterior_mean", "posterior_mode")

# Report strategy at every reference point:
# - component-local unscaled ESS
# - weighted hard-assignment ESS (paper definition)
local_ess_row <- function(th) {
  w_mix <- as.numeric(mix$weights)
  w_mix <- w_mix / sum(w_mix)

  # This vignette is 1D, so we can evaluate component responsibilities directly.
  log_comp <- vapply(seq_along(w_mix), function(j) {
    mu_j <- as.numeric(mix$mu[[j]])
    sd_j <- sqrt(as.numeric(mix$Sigma[[j]]))
    log(w_mix[j]) + dnorm(th, mean = mu_j, sd = sd_j, log = TRUE)
  }, numeric(1))
  resp <- exp(log_comp - logsumexp(log_comp))
  k_star <- which.max(resp)

  ess_components <- vapply(seq_along(w_mix), function(j) {
    Ipi_j <- if (!is.null(mix$prec) && length(mix$prec) >= j && !is.null(mix$prec[[j]])) {
      as.matrix(mix$prec[[j]])
    } else {
      solve(mix$Sigma[[j]])
    }
    as.numeric(geoess_trace(I1, Ipi_j))
  }, numeric(1))

  tol <- 1e-10
  tie_idx <- which(abs(resp - max(resp)) <= tol)
  ess_hard <- mean(w_mix[tie_idx] * ess_components[tie_idx])

  weight_str <- if (length(tie_idx) == 1L) {
    sprintf("%.3f", w_mix[k_star])
  } else {
    paste(sprintf("%.3f", w_mix[tie_idx]), collapse = " / ")
  }

  ess_comp_str <- if (length(tie_idx) == 1L) {
    sprintf("%.1f", ess_components[k_star])
  } else {
    paste(sprintf("%.1f", ess_components[tie_idx]), collapse = " / ")
  }

  list(
    Weights = weight_str,
    ESS_component_unscaled = ess_comp_str,
    ESS_local = as.numeric(ess_hard)
  )
}

ess_rows <- lapply(ref_points, function(th) {
  local_ess_row(th)
})

local_table <- data.frame(
  Weights = vapply(ess_rows, `[[`, character(1), "Weights"),
  ESS_component_unscaled = vapply(ess_rows, `[[`, character(1), "ESS_component_unscaled"),
  ESS_local = round(vapply(ess_rows, `[[`, numeric(1), "ESS_local"), 1),
  check.names = FALSE
)
rownames(local_table) <- ref_labels
invisible(local_table)
```

## Plot: sample-based prior and reference points

```{r prior-plot, fig.width=14, fig.height=5, fig.cap="Reference points used in the bimodal-prior example."}
d_prior <- density(prior_draws)
plot(d_prior, main = "Reference points in the bimodal-prior example",
     xlab = expression(theta), ylab = "density")
abline(v = prior_mean, col = "steelblue", lwd = 2)
abline(v = prior_modes, col = "darkgreen", lwd = 2, lty = 2)
abline(v = prior_misspec, col = "firebrick", lwd = 2, lty = 3)
abline(v = post_mean, col = "purple", lwd = 2)
abline(v = post_mode, col = "orange", lwd = 2, lty = 2)

legend("topright",
       legend = c("prior mean", "prior modes", "prior misspec",
                  "posterior mean", "posterior mode"),
       col = c("steelblue", "darkgreen", "firebrick", "purple", "orange"),
       lwd = 2, lty = c(1, 2, 3, 1, 2), bty = "n")
```

## Global ESS (prior and posterior averages)

Prior-averaged ESS uses the prior draws directly. Posterior-averaged ESS is
approximated by reweighting those draws with the likelihood. For comparison, we
also retain the older full-mixture global ESS calculation as an unweighted
reference summary.

```{r global-ess}
has_global_attr <- "mixture_attribution" %in% names(formals(geoess_global))

if (has_global_attr) {
  ESS_global_prior <- geoess_global(
    Q = matrix(prior_draws, ncol = 1),
    I1 = I1,
    Ipi = mix,
    method = "trace",
    mixture_attribution = "hard_resp",
    psd = "clip"
  )

  ESS_global_post <- geoess_global(
    Q = matrix(prior_draws, ncol = 1),
    I1 = I1,
    Ipi = mix,
    method = "trace",
    mixture_attribution = "hard_resp",
    psd = "clip",
    weights = post_w
  )

  ESS_global_prior_unweighted <- geoess_global(
    Q = matrix(prior_draws, ncol = 1),
    I1 = I1,
    Ipi = mix,
    method = "trace",
    mixture_attribution = "mix",
    psd = "clip"
  )

  ESS_global_post_unweighted <- geoess_global(
    Q = matrix(prior_draws, ncol = 1),
    I1 = I1,
    Ipi = mix,
    method = "trace",
    mixture_attribution = "mix",
    psd = "clip",
    weights = post_w
  )
} else {
  ess_local_all <- vapply(prior_draws, function(th) local_ess_row(th)$ESS_local, numeric(1))
  post_w_norm <- post_w / sum(post_w)

  ESS_global_prior <- list(estimate = mean(ess_local_all))
  ESS_global_post <- list(estimate = sum(post_w_norm * ess_local_all))

  ESS_global_prior_unweighted <- geoess_global(
    Q = matrix(prior_draws, ncol = 1),
    I1 = I1,
    Ipi = mix,
    method = "trace",
    psd = "clip"
  )

  ESS_global_post_unweighted <- geoess_global(
    Q = matrix(prior_draws, ncol = 1),
    I1 = I1,
    Ipi = mix,
    method = "trace",
    psd = "clip",
    weights = post_w
  )
}

global_table <- data.frame(
  prior_avg = ESS_global_prior$estimate,
  posterior_avg = ESS_global_post$estimate,
  unweighted_prior_avg = ESS_global_prior_unweighted$estimate,
  unweighted_posterior_avg = ESS_global_post_unweighted$estimate
)
invisible(global_table)
```

## RBesT comparison (prior-only)

We fit a normal mixture to the same draws using RBesT and compute
ELIR / moment / Morita ESS as a baseline comparison.

```{r rbest-compare}
rbest_vals <- if (requireNamespace("RBesT", quietly = TRUE)) {
  rbest_ess_from_draws(prior_draws, sigma = sigma)
} else {
  c(elir = NA_real_, moment = NA_real_, morita = NA_real_)
}
invisible(rbest_vals)
```

## ESS Comparison Summary

```{r ess-comparison-display}
format_slash_round <- function(x, digits = 1) {
  parts <- trimws(strsplit(as.character(x), "/", fixed = TRUE)[[1]])
  vals <- suppressWarnings(as.numeric(parts))
  if (all(is.finite(vals))) {
    paste(format(round(vals, digits), nsmall = digits, trim = TRUE), collapse = " / ")
  } else {
    x
  }
}

local_panel <- data.frame(
  `Reference Point` = c(
    "Prior Mean",
    sprintf("Prior Mode near %.2f", ref_points[2]),
    sprintf("Prior Mode near %.2f", ref_points[3]),
    "Saddle point",
    "Posterior Mean",
    "Posterior Mode"
  ),
  `Reference \\(\\theta^*\\)` = round(ref_points, 3),
  Weights = vapply(local_table$Weights, format_slash_round, character(1), digits = 1),
  `Unweighted value` = vapply(local_table$ESS_component_unscaled, format_slash_round, character(1), digits = 1),
  Value = round(local_table$ESS_local, 1),
  check.names = FALSE
)

global_panel <- data.frame(
  `Q-averaged` = c("Prior-averaged", "Posterior-averaged"),
  Value = round(c(global_table$prior_avg, global_table$posterior_avg), 1),
  `Unweighted value` = round(c(global_table$unweighted_prior_avg, global_table$unweighted_posterior_avg), 1),
  check.names = FALSE
)

existing_panel <- data.frame(
  Method = c("Morita", "Moment", "ELIR"),
  Value = round(c(rbest_vals["morita"], rbest_vals["moment"], rbest_vals["elir"]), 1),
  check.names = FALSE
)

knitr::kable(
  local_panel,
  escape = FALSE,
  caption = paste(
    "Panel A: local ESS for the bimodal prior example.",
    "The weights and unweighted values refer to the local component;",
    "at the saddle point, both tied components are shown."
  )
)

knitr::kable(
  global_panel,
  escape = FALSE,
  caption = "Panel B: global ESS for the bimodal prior example."
)

knitr::kable(
  existing_panel,
  escape = FALSE,
  caption = "Panel C: existing ESS estimated from RBesT for the bimodal prior example."
)
```

Panel A shows that local ESS varies substantially across reference points. The
values range from 47.9 to 63.0 depending on the location of the reference
point. The prior mean and the mode near 0.87 both yield ESS 63.0, reflecting
the same dominant Gaussian component. By contrast, the mode near 0.10, the
posterior mean, and the posterior mode all yield ESS 47.9, because those
reference points are attributed to the other local component. At the saddle
point, the weighted tie rule in Equation (1) gives an
intermediate value of 55.5. This is exactly the kind of reference-point
dependence that a single ESS number would miss.

Panel B reports the global ESS summaries. The prior-averaged and
posterior-averaged values are 58.6 and 54.0, respectively, when the weighted
hard-assignment local ESS is used. For comparison, the older unweighted
full-mixture summaries are much larger at 110.4 and 131.9. This contrast shows
that the choice of local attribution rule carries through directly into the
global summary.

Panel C compares the same prior with existing sample-based ESS methods from
`RBesT`. Morita ESS is 88.6, close to the unweighted local value 89.5 from the
dominant Gaussian component. The moment method gives only 8.2, whereas ELIR is
110.7, close to the unweighted prior-averaged global ESS 110.4. These large
discrepancies illustrate that existing methods can behave quite differently for
multimodal priors, especially when they implicitly collapse the geometry to a
single-number summary.

In this vignette:

- `ESS_component_unscaled` is the unscaled ESS of the max-responsibility component.
- `ESS_local` is the paper definition based on weighted hard assignment:
  \(w_{k^*}\,ESS_{k^*}(\theta^*)\) for a unique winner, and
  \(\frac{1}{|T|}\sum_{k\in T} w_k\,ESS_k(\theta^*)\) when responsibilities tie.
  In particular, at a tied saddle point between two components, we use
  (1), that is, the average of the tied components' weighted ESS
  values.

Global ESS averages local ESS values under an explicitly chosen weighting
distribution. For mixture priors, the default global summary in
`geoess_global()` uses the same weighted hard-assignment rule as the local ESS
definition.

Taken together, the local and global summaries show why multiple ESS should be
reported for complex priors. Local ESS describes how informative the prior is
at a specific reference point, whereas global ESS describes average
informativeness under a stated weighting distribution. For bimodal or otherwise
complex priors, both are needed to characterize prior influence adequately.
